Safety Cases for Software Product Lines: Terminology and Example

### 2017-10-17 doc/SafetyCases/SafetyCasesPL.tex
minor document updates

\title{Safety Cases for Software Product Lines}
%\author{The Author}
\date{}                                           % Activate to display a given date or no date

@@

\section{Introduction}
Software systems are everywhere, including safety critical domains. Software is inherently embedded into medical devices, power plant controllers, airplane controllers, drones, and soon self-driving vehicles. Safety analysis of this category of systems is an essential ingredient of the Software Development Lifecycle (SDLC). Safety cases are the most common artifacts resulting from safety analysis.

Orthogonally, most of the aforementioned examples of safety critical software systems are engineered as Software Product Lines (SPLs). Variability is inherent in families of software systems. Variability can be due to different deployment environments, differences in hardware devices and peripherals, varying requirements across models of systems, and difference in performance characteristics.

Variability in an SPL is usually measured in terms of the number of features in a product line. The number of possible products that can be synthesized from an SPL grows exponentially with the number of features. As a result, enumerating all possible products and performing safety analysis at the product level is intractable.

Several attempts have been made to leverage the high degree of commonality across products of an SPL to minimize the cost of product line analysis in general, and safety analysis in particular. This document explores previous attempts and fundamental questions at the crossroads of Safety Analysis and Software Product Lines.

\subsection{Goals}
This document aims at:
\begin{itemize}
\item Introducing terminology for both Software Product Line and Safety Case concepts
\item Demonstrating the Product Line and Safety Case concepts on a simple exemplar
\item To come to a mutual agreement on concepts, terminology and the Product Line safety engineering at GM
\end{itemize}

@@

\begin{enumerate}
\item Displaying the fuel level of a fuel tank on the driver's dashboard
\item Triggering a warning when the fuel level goes below a certain threshold (optional feature)
\end{enumerate}

@@

\begin{enumerate}
\item Truck with liquid fuel
\item Bus with liquid fuel
\item Truck with gas (doesn't include the fuel level warning feature)
\item Bus with gas
\end{enumerate}

@@

\begin{enumerate}
\item a Kalman filter is used in trucks
\item a Low-Pass filter is used in buses
\end{enumerate}

@@

Variability is commonplace in software systems. Managing variability across different software artifacts, and throughout the Software Development Life Cycle (SDLC) is always a challenge though. Software Product Line Engineering draws an analogy to industrial product lines, where different variants of a product are manufactured using a common set of parts and machinery. The goal of SPLs is to maximize reuse across different variants of a software product, while at the same time simplifying the extensibility and maintainability of the product line.

@@

For our running example, we have four variants of Fuel Level Display systems. In addition, each of those variants can come with or without the warning feature. This family of product variants has a lot in common, so it is more cost effective to build them as an SPL instead of building each product in isolation.

@@

Orthogonally, variation with respect to vehicle types and fuel types are also considered feature variants.

So this is the FLDPL list of features:

@@

$(Truck \vee Bus) \wedge (\neg Truck \vee \neg Bus) $

@@

\begin{figure}
  \centering
    \includegraphics[width=0.75\textwidth]{FeatureDiagram}
  \caption{Feature Diagram Example}
  \label{fig:FeatureDiagram}
\end{figure}

For example, figure \ref{fig:FeatureDiagram} shows a feature diagram for our SPL.

@@

@@

\begin{lstlisting}[language=Java, caption=Generative example, label={lst:Generative}]
// Feature module FLDGas
class FLD {
	void reportFuelLevel(float level) { 
		dashboard.display(level); 
	}
}

@@

Listing \ref{lst:Generative} shows a Feature-Oriented high-level implementation of FLDPL. \emph{FLDGas}, \emph{FLDLiquidTruck} and \emph{FLDLiquidBus} are three separate implementations of the Fuel Level Display module. \emph{LowGasWarning} is implemented as a cross-cutting feature that refines any of the 3 \emph{FLD} implementations.

\begin{lstlisting}[language=Java, caption=Generated Product, label={lst:Generated}]
class FLD {
	const float THRESHOLD;

@@

Given the feature expression (Liquid $\wedge$ Truck $\wedge$ FuelWarning), the product in listing \ref{lst:Generated} is generated.

@@

If we are to rewrite FLDPL using CPP macros, it would look like listing 
ref{lst:Annotative}. A single feature (e.g. \emph{LOW\_GAS\_WARNING} is now scattered across the code base, and in some cases replicated.

\begin{lstlisting}[language=Java, caption=Annotative example, label={lst:Annotative}]
class FLD {
	#if defined (LOW_GAS_WARNING)
	const float THRESHOLD;
	#endif

@@

\begin{lstlisting}[caption=Delta-Oriented example, label={lst:Delta}]
core FLDGas {
	class FLD {
		void reportFuelLevel(float level) { 
			dashboard.display(level); 
		}
	} 
}

@@

Goal Structuring Notation is used to layout logical arguments connecting safety goals to pieces of evidence satisfying them. A safety case in GSN is typically composed of 3 kinds of artifacts:

\begin{figure}
  \centering
  \includegraphics[width=0.75\textwidth]{gsn}
  \caption{GSN Elements}
\end{figure}

@@

@@

\section{Product Line Engineering Scenarios}

Software Product Lines can come into existence in several ways. This section briefly outline engineering scenarios for building Product Lines.

\subsection{Product Evolution over Time}

Starting with a single product, variability might start exhibiting itself in lots of different ways. Different deployment environments, different modes of operations, minor requirement variations from different stakeholders, and varying workloads are all examples of product-level variability. Over time, the maintenance of software artifacts with too much ad-hoc variability becomes a real problem.

One way to manage ad-hoc variability is to evolve a single product with variability in a product line. This process usually starts with factoring out a common core (common to all variants), and variant-specific artifacts.

\subsection{Product Component Reuse}

Again starting with a single component-oriented product, some components might be potentially reused in other products. Sometimes a component can be reused as is, without the need for any modification. In other cases the component might require minor modifications, resulting in a slightly different version.

This component-level variability can again evolve into a component-level Product Line. With more and more components being reused in this fashion, a component Product Line library becomes the skeleton of a Software Product Line.

\subsection{Explicit Product-Line Engineering}

The previously outlined scenarios start with a single product, and with more and more variability introduced the product evolves into a product line. Sometimes on the other hand the designers of a system are aware from day one that they are building a family or products, not just one. If at design time they draw a clear boundary between a common core and a set of varying artifacts, they are actually engineering a Product Line.

%\subsection{Product Lines and Safety Cases}
%
%Reasoning about Software Product Lines is not straightforward because of several reasons:
%
%\begin{itemize}
%
%\item The number of possible valid products of a product lines grows exponentially with the number of features. As a result, enumerating all possible valid products becomes intractable, and thus existing product-level reasoning techniques don't directly scale to product lines.
%
%\item A Feature Model is an integral part of a Product Line, and its role is to define the valid set of feature combinations. When reasoning about a Product Line, the Feature Model has to be taken into consideration to avoid including invalid products in any argument.
%
%\item Whenever an argument is made regarding a Product Line, it is essential to specify whether this argument applies to all valid products (universal quantification over products), or only to some (existential quantification). 
%
%\end{itemize}
%
%\subsection{Q1: Product-Line of Safety Cases or a Safety Case of a Product Line}
%
%Safety Analysis is one example of reasoning about a software systems. Techniques like Hazard Analysis, Fault Tree Analysis and Safety Cases have been well studied and applied to software products. However, when it comes to Product Lines, those techniques do not directly scale due to the previously mentioned reasons. 
%
%In addition, one fundamental question pertains to the granularity to which Safety Analysis is performed. We can think of 2 different approaches here:
%
%\begin{itemize}
%
%\item A Safety Case of a Product Line: here safety analysis as a process is applied to a Software Product Line, resulting in a single Safety Case covering the whole Product Line.
%
%\item A Product Line of Safety Cases: since a Safety Case itself is a Software Artifact (albeit a semi-formal one at best), we can have a Product Line of Safety Cases (pretty much like Product Lines of Use Case Diagrams, Class Diagrams or Source Code artifacts). 
%
%\end{itemize}
%
%\subsection{Product Line evolution impact analysis}
%
%As an SPL evolves over time (with changing requirements or more variability), existing safety cases need to evolve as well. This is essential to maintain the validity of the safety arguments. However, the safety assurance process is expensive in terms of resources and is also time consuming and labor intensive. As a result, as much of existing evidences should be reused as possible, without jeopardizing the validity of the safety arguments. Two questions are relevant here:
%
%\begin{itemize}
%
%\item Q2: As an SPL evolves, how can we determine which pieces of evidence are still valid, and can be reused as is, as opposed to pieces of evidence that are invalidated by the SPL changes?
%
%\item Q3: Within a single piece of evidence, how much reuse can be achieved? For example, if the results of a test suite and considered a piece of evidence, how much of the test cases are common across variants and how much are variant-specific? How easy is it to draw a correspondence between features, goals/sub-goals and evidences/sub-evidences? 
%
%\end{itemize}

>>>>

\title{Safety Cases for Software Product Lines: Terminology and Example}
\author{Ramy Shahin \\ University of Toronto \\ rshahin@cs.toronto.edu \and
Marsha Chechik \\ University of Toronto \\ chechik@cs.toronto.edu}
\date{\today}                                           % Activate to display a given date or no date

@@

\section{Introduction}
%Software systems are everywhere, including safety critical domains. Software is inherently embedded into medical devices, power plant controllers, airplane controllers, drones, and soon self-driving vehicles. Safety analysis of this category of systems is an essential ingredient of the Software Development Lifecycle (SDLC). Safety cases are the most common artifacts resulting from safety analysis.

%Orthogonally, most of the aforementioned examples of safety critical software systems are engineered as Software Product Lines (SPLs). Variability is inherent in families of software systems. Variability can be due to different deployment environments, differences in hardware devices and peripherals, varying requirements across models of systems, and difference in performance characteristics.

%Variability in an SPL is usually measured in terms of the number of features in a product line. The number of possible products that can be synthesized from an SPL grows exponentially with the number of features. As a result, enumerating all possible products and performing safety analysis at the product level is intractable.

Several attempts have been made to leverage the high degree of commonality across products of a Software Product Line (SPL) to minimize the cost of product line analysis in general, and safety analysis in particular. This document explores previous attempts and fundamental questions at the crossroads of Safety Analysis and Software Product Lines.

\subsection{Goals}
This document aims at:
\begin{itemize}
\item Introducing terminology for both Software Product Line and Safety Case concepts
\item Demonstrating the Product Line and Safety Case concepts on a simple exemplar
\item To try to come to a mutual agreement on concepts, terminology and the Product Line safety engineering at GM
\end{itemize}

@@

\begin{itemize}
\item Displaying the fuel level of a fuel tank on the driver's dashboard
\item Triggering a warning when the fuel level goes below a certain threshold (optional feature)
\end{itemize}

@@

\begin{itemize}
\item Truck with liquid fuel
\item Bus with liquid fuel
\item Truck with gas (doesn't include the fuel level warning feature)
\item Bus with gas
\end{itemize}

@@

\begin{itemize}
\item a Kalman filter is used in trucks
\item a Low-Pass filter is used in buses
\end{itemize}

@@

%Variability is commonplace in software systems. Managing variability across different software artifacts, and throughout the Software Development Life Cycle (SDLC) is always a challenge though. Software Product Line Engineering draws an analogy to industrial product lines, where different variants of a product are manufactured using a common set of parts and machinery.

The goal of SPLs is to maximize reuse across different variants of a software product, while at the same time simplifying the extensibility and maintainability of the product line.

@@

For our running example, we have four variants of Fuel Level Display systems. In addition, each of those variants can come with or without the warning feature. This family of product variants has a lot in common, so it makes sense to build them as an SPL instead of building each product in isolation.

@@

Orthogonally, there is variation with respect to vehicle types and fuel types.

The FLDPL list of features is:

@@

(Truck $\vee$ Bus) $\wedge$ ($\neg$ Truck $\vee \neg$ Bus)

@@

\begin{figure}
  \centering
    \includegraphics[width=0.75\textwidth]{FeatureDiagram}
  \caption{Feature Diagram of the FLDPL}
  \label{fig:FeatureDiagram}
\end{figure}

For example, figure \ref{fig:FeatureDiagram} shows a feature diagram of the FLDPL. In this diagram, the usage context is composed of two mandatory features: Vehicle type and Fuel type. Vehicle type can be either Truck or Bus (mutually exclusive options). Similarly, Fuel type can be either Liquid or Gas. In addition, the FLD system has one mandatory feature (Fuel Level Estimation) and one optional feature (Low Fuel Level Warning). An annotation (O1) is added to the diagram to specify the set of products where the low fuel level warning feature is present.

@@

\subsection{Product Configuration}

\item[Product Configuration]
A product configuration is a boolean formula over the set of features of an SPL that specifies the set of features present in a product.
\end{description}

For example, the configuration of an FLD for a Truck using Gas as fuel would be (Truck $\wedge$ Gas $\wedge$ FuelLevel).

\subsection{150\% Representation}

\item[150\% Representation]
The 150\% representation of an SPL is a syntactic representation with syntactic constructs belonging to all the features of the SPL. 
\end{description}

Since in the vast majority of cases all those syntactic constructs can not all coexist in a product (given feature model constraints), this representation is referred to as 150\%, as opposed to the 100\% representation of a single product.

Examples of 150\% representations can be found in section 4.

@@

\begin{lstlisting}[language=Java, caption=Generative FLDPL, label={lst:Generative}, float]
// Feature module FLDGas
class FLD {
	void reportFuelLevel(float level) { 
		dashboard.display(level); 
	}
}

@@

Listing \ref{lst:Generative} shows a Feature-Oriented high-level implementation of FLDPL. FLDGas, FLDLiquidTruck and FLDLiquidBus are three separate implementations of the Fuel Level Display module. LowGasWarning is implemented as an orthogonal feature that refines any of the 3 FLD implementations.

\begin{lstlisting}[language=Java, caption=Generated FLD Product, label={lst:Generated}, float]
class FLD {
	const float THRESHOLD;

@@

Given the product configuration (Liquid $\wedge$ Truck $\wedge$ FuelWarning), the product in listing \ref{lst:Generated} is generated.

@@

If we are to rewrite FLDPL using CPP macros, it would look like listing \ref{lst:Annotative}. A single feature implementation (e.g. LOW\_GAS\_WARNING is now scattered across the code base, and in some cases replicated. This code snippet is an example of a 150\% representation, where a product configuration (in terms of macro definitions) excludes some of the syntactic constructs, leaving only those belonging to a specific product.

\begin{lstlisting}[language=Java, caption=Annotative FLDPL, label={lst:Annotative}, float]
class FLD {
	#if defined (LOW_GAS_WARNING)
	const float THRESHOLD;
	#endif

@@

\begin{lstlisting}[caption=Delta-Oriented FLDPL, label={lst:Delta}, float]
core FLDGas {
	class FLD {
		void reportFuelLevel(float level) { 
			dashboard.display(level); 
		}
	} 
}

@@

Goal Structuring Notation \cite{Kelly} can be used to layout logical arguments connecting safety goals to pieces of evidence satisfying them. As seen in figure \ref{fig:gsn}, a safety case in GSN is typically composed of 3 kinds of artifacts:

\begin{figure}
  \centering
  \includegraphics[width=0.75\textwidth]{gsn}
  \caption{GSN Elements}
  \label{fig:gsn}
\end{figure}

@@

Here the robustness of the estimation algorithm in the context of using the Kalman filter is a goal (G\_6). This goal is refined into a more specific goal: the steadiness of the output of the Kalman filter (G\_7).  This goal is further refined into 3 sub-goals (G\_8, G\_9, G\_10). Goal G\_8 s fulfilled by the results of standard deviation analysis (evidence E\_1). In addition, goals G\_9 and G\_10 are fulfilled by simulation results (evidence E\_2).

@@

\section{Questions}

To make sure all parties agree on the context of this project, and that research work is aligned with GM practical problems, we have a few questions:

\begin{enumerate}

\item Is the example used in this document (FLDPL) representative of the SPLs developed at GM?

\item What is the product line engineering approach used by GM engineers (annotative, generative or delta-oriented)?

\item What is the safety case notation used by GM safety engineers?

\item How is safety case variability currently maintained by GM safety engineers?

\item How are hazards, safety case goals, pieces of evidence managed across different products in a product line by GM safety engineers?

\end{enumerate}

%\section{Product Line Engineering Scenarios}
%
%Software Product Lines can come into existence in several ways. This section briefly outline engineering scenarios for building Product Lines.
%
%\subsection{Product Evolution over Time}
%
%Starting with a single product, variability might start exhibiting itself in lots of different ways. Different deployment environments, different modes of operations, minor requirement variations from different stakeholders, and varying workloads are all examples of product-level variability. Over time, the maintenance of software artifacts with too much ad-hoc variability becomes a real problem. 
%
%One way to manage ad-hoc variability is to evolve a single product with variability in a product line. This process usually starts with factoring out a common core (common to all variants), and variant-specific artifacts.
%
%\subsection{Product Component Reuse}
%
%Again starting with a single component-oriented product, some components might be potentially reused in other products. Sometimes a component can be reused as is, without the need for any modification. In other cases the component might require minor modifications, resulting in a slightly different version. 
%
%This component-level variability can again evolve into a component-level Product Line. With more and more components being reused in this fashion, a component Product Line library becomes the skeleton of a Software Product Line.
%
%\subsection{Explicit Product-Line Engineering}
%
%The previously outlined scenarios start with a single product, and with more and more variability introduced the product evolves into a product line. Sometimes on the other hand the designers of a system are aware from day one that they are building a family or products, not just one. If at design time they draw a clear boundary between a common core and a set of varying artifacts, they are actually engineering a Product Line.
%
%\subsection{Product Lines and Safety Cases}
%
%Reasoning about Software Product Lines is not straightforward because of several reasons:
%
%\begin{itemize}
%
%\item The number of possible valid products of a product lines grows exponentially with the number of features. As a result, enumerating all possible valid products becomes intractable, and thus existing product-level reasoning techniques don't directly scale to product lines.
%
%\item A Feature Model is an integral part of a Product Line, and its role is to define the valid set of feature combinations. When reasoning about a Product Line, the Feature Model has to be taken into consideration to avoid including invalid products in any argument.
%
%\item Whenever an argument is made regarding a Product Line, it is essential to specify whether this argument applies to all valid products (universal quantification over products), or only to some (existential quantification). 
%
%\end{itemize}
%
%\subsection{Q1: Product-Line of Safety Cases or a Safety Case of a Product Line}
%
%Safety Analysis is one example of reasoning about a software systems. Techniques like Hazard Analysis, Fault Tree Analysis and Safety Cases have been well studied and applied to software products. However, when it comes to Product Lines, those techniques do not directly scale due to the previously mentioned reasons. 
%
%In addition, one fundamental question pertains to the granularity to which Safety Analysis is performed. We can think of 2 different approaches here:
%
%\begin{itemize}
%
%\item A Safety Case of a Product Line: here safety analysis as a process is applied to a Software Product Line, resulting in a single Safety Case covering the whole Product Line.
%
%\item A Product Line of Safety Cases: since a Safety Case itself is a Software Artifact (albeit a semi-formal one at best), we can have a Product Line of Safety Cases (pretty much like Product Lines of Use Case Diagrams, Class Diagrams or Source Code artifacts). 
%
%\end{itemize}
%
%\subsection{Product Line evolution impact analysis}
%
%As an SPL evolves over time (with changing requirements or more variability), existing safety cases need to evolve as well. This is essential to maintain the validity of the safety arguments. However, the safety assurance process is expensive in terms of resources and is also time consuming and labor intensive. As a result, as much of existing evidences should be reused as possible, without jeopardizing the validity of the safety arguments. Two questions are relevant here:
%
%\begin{itemize}
%
%\item Q2: As an SPL evolves, how can we determine which pieces of evidence are still valid, and can be reused as is, as opposed to pieces of evidence that are invalidated by the SPL changes?
%
%\item Q3: Within a single piece of evidence, how much reuse can be achieved? For example, if the results of a test suite and considered a piece of evidence, how much of the test cases are common across variants and how much are variant-specific? How easy is it to draw a correspondence between features, goals/sub-goals and evidences/sub-evidences? 
%
%\end{itemize}
